---
title: "Lab 1 Script"
author: "Robert Deng, Tiffany Jaya, Shan He, Joanna Huang"
output: 
  pdf_document:
    latex_engine: xelatex
  toc: true
  number_sections: true
fontsize: 11pt
geometry: margin=1in
---

```{r}
# Code 1
summary(challenger)
describe(challenger)
ggplot(challenger, aes(x = Temp)) + geom_histogram(binwidth = 3)

# list rows of data that have missing values 
challenger[!complete.cases(challenger),] # no missing values

```

```{r}
# Code 2
ggplot(challenger, aes(x = Temp, y=O.ring)) + geom_point() + 
  labs(title = 'Joint Temperature Versus Number of O-Ring Failures', y='Number of O-ring failures', x='Temp(F)') + 
  theme(plot.title = element_text(lineheight=1, face="bold", color = "darkred"))
```

```{r}
# Code 3
ggplot(challenger, aes(x=Pressure, y=O.ring)) + geom_point() + 
  labs(title = 'Joint Pressure Versus Number of O-Ring Failures', y='Number of O-ring failures', x='Pressure (psi') + 
  theme(plot.title = element_text(lineheight=1, face="bold", color = "lightseagreen")) 
  
challenger %>%
  ggplot(aes(x = factor(O.ring), fill = factor(O.ring))) +
  geom_bar() +
  facet_wrap(~Pressure) + 
  labs(title = 'Count of O-Ring Failures at Pressure Levels', y='Count', x='Number of O-ring failures') + 
  theme(plot.title = element_text(lineheight=1, face="bold", color = "lightseagreen")) 
```

```{r}
# Code 4
challenger$Fail <- ifelse(challenger$O.ring>0, 1, 0)
head(challenger)

```
```{r}
# Notation code - Code 5
```
We apply the logistic regression model to model the probability of the O-rings failing at various temperatures.

$$\begin{aligned} P(Y_i = 1 | x_i) = \frac{e^{\beta_0 + \beta_1 x_i}}{1 + e^{\beta_0 + \beta_1 x_i}} = \pi_i \end{aligned}$$

$i$: observation for a single launch
$Y_i = 1$: failure launch, at least one O-ring failing
$Y_i = 0$: successful launch, no O-ring failing
$x_i$: outside temperature in degrees Fahrenheit

```{r}
# Notation code - Code 6
```
$H_0: \beta_1 = 0$

$H_a: \beta_1 \ne 0$

Null Hypothesis: outside temperature has no effect in O-rings failure

Alternative Hypothesis: outside temperature has an effect in O-rings failure

```{r}
# Code 7
m0 <- glm(formula = Fail ~ 1, 
          family = binomial,
          data = challenger)
m1 <- glm(formula = Fail ~ Temp, 
          family = binomial,
          data = challenger)
summary(m1)
m1.b0 <- m1$coefficients[1]
m1.b1 <- m1$coefficients[2]
```

```{r}
# code 8
lp.31 <- predict(object = m1, newdata= data.frame(Temp=31), type = 'link', se = TRUE)

lp.31.pred <- lp.31$fit 
lp.31.lin.pred.lower <- lp.31$fit - qnorm(p = 1-0.05/2) * lp.31$se
lp.31.lin.pred.upper <- lp.31$fit + qnorm(p = 1-0.05/2) * lp.31$se

pi.31.pred <- exp(lp.31$fit) / (1+exp(lp.31$fit))
pi.31.lower <- exp(lp.31.lin.pred.lower) / (1+exp(lp.31.lin.pred.lower))
pi.31.upper <- exp(lp.31.lin.pred.upper) / (1+exp(lp.31.lin.pred.upper))

pi.31.pred
pi.31.lower
pi.31.upper
```

```{r}
# Code 9
# bootstrap for 5(e)

bootstrap_coef <- function(data, model, pred_temp) {
  #sample temp with replacement
  sample_temp <- sample(x = data, size = 23, replace = TRUE)
  #find predicted probability
  sample_preds <-predict(object = model, newdata = data.frame(Temp = sample_temp), type = "response")

  #build a dataframe
  df <- data.frame(temp = sample_temp, 
             prob_pred = sample_preds
             )
  
  #map function to find predicted class (fail or not fail)
  df$fail_pred <- unlist(lapply(X = df$prob_pred, FUN = function(x) rbinom(1, 1, x)))
  
  #build glm model
  bootstrap_glm <- glm(formula = fail_pred ~ temp, family = binomial(link=logit), data = df)
  
  #predict probability at pred_temp
  predict(object = bootstrap_glm, newdata = data.frame(temp = pred_temp), type = "response")
}

est_prob_31 <- replicate(10000, bootstrap_coef(challenger$Temp, m1, 31), simplify = FALSE)

est_prob_72 <- replicate(10000, bootstrap_coef(challenger$Temp, m1, 72), simplify = FALSE)

quantile(unlist(est_prob_31), c(.05, .95))
quantile(unlist(est_prob_72), c(.05, .95))
```

```{r}
# Code 10
m4 <- glm(formula = Fail ~ Temp + I(Temp^2), family = binomial(link=logit), data = challenger) 
summary(m4)
anova(m1, m4, test='Chisq')
```

