---
title: "Lab 1 Script"
author: "Robert Deng, Tiffany Jaya, Shan He, Joanna Huang"
output: 
  pdf_document:
    latex_engine: xelatex
  toc: true
  number_sections: true
fontsize: 11pt
geometry: margin=1in
---

```{r}
# Code 1
summary(challenger)
describe(challenger)
ggplot(challenger, aes(x = Temp)) + geom_histogram(binwidth = 3)

# list rows of data that have missing values 
challenger[!complete.cases(challenger),] # no missing values

```

```{r}
# Code 2
ggplot(challenger, aes(x = Temp, y=O.ring)) + geom_point() + 
  labs(title = 'Joint Temperature Versus Number of O-Ring Failures', y='Number of O-ring failures', x='Temp(F)') + 
  theme(plot.title = element_text(lineheight=1, face="bold", color = "darkred"))
```

```{r}
# Code 3
ggplot(challenger, aes(x=Pressure, y=O.ring)) + geom_point() + 
  labs(title = 'Joint Pressure Versus Number of O-Ring Failures', y='Number of O-ring failures', x='Pressure (psi') + 
  theme(plot.title = element_text(lineheight=1, face="bold", color = "lightseagreen")) 
  
challenger %>%
  ggplot(aes(x = factor(O.ring), fill = factor(O.ring))) +
  geom_bar() +
  facet_wrap(~Pressure) + 
  labs(title = 'Count of O-Ring Failures at Pressure Levels', y='Count', x='Number of O-ring failures') + 
  theme(plot.title = element_text(lineheight=1, face="bold", color = "lightseagreen")) 
```

```{r}
# Code 4
challenger$Fail <- ifelse(challenger$O.ring>0, 1, 0)
head(challenger)

```
```{r}
# Notation code - Code 5
```
We apply the logistic regression model to model the probability of the O-rings failing at various temperatures.

$$\begin{aligned} P(Y_i = 1 | x_i) = \frac{e^{\beta_0 + \beta_1 x_i}}{1 + e^{\beta_0 + \beta_1 x_i}} = \pi_i \end{aligned}$$

$i$: observation for a single launch
$Y_i = 1$: failure launch, at least one O-ring failing
$Y_i = 0$: successful launch, no O-ring failing
$x_i$: outside temperature in degrees Fahrenheit

```{r}
# Notation code - Code 6
```
$H_0: \beta_1 = 0$

$H_a: \beta_1 \ne 0$

Null Hypothesis: outside temperature has no effect in O-rings failure

Alternative Hypothesis: outside temperature has an effect in O-rings failure

```{r}
Code 7
m0 <- glm(formula = Fail ~ 1, 
          family = binomial,
          data = challenger)
m1 <- glm(formula = Fail ~ Temp, 
          family = binomial,
          data = challenger)
summary(m1)
m1.b0 <- m1$coefficients[1]
m1.b1 <- m1$coefficients[2]
```
