---
title: "Lab 1 Script"
author: "Robert Deng, Tiffany Jaya, Shan He, Joanna Huang"
output: 
  pdf_document:
    latex_engine: xelatex
  toc: true
  number_sections: true
fontsize: 11pt
geometry: margin=1in
---

# Code 0: load packages and dataset

```{r}
# add packages
library(dplyr)
library(ggplot2)
library(knitr)
library(stargazer)
# prevent source code from running off the page
opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
# remove all objects from current workspace
rm(list = ls())
# load data
d <- read.csv("./dataset/challenger.csv", header=TRUE, sep=",")
# set x-axis min and max for temp and temp at launch
xlim.min <- 30; xlim.max <- 80; temp.launch <- 31
```

# Code 1A: EDA on the structure of the data

```{r}
# structure of data
str(d)
# summary of data
summary(d)
# list row with missing values
d[!complete.cases(d),]
```

# Code 1B: Univariate analysis on O.ring

```{r}
# number of O-rings failure
table(d$O.ring)
```

```{r}
hist(d$O.ring, xlab='Number of O-rings failure', breaks=0:3 - 0.5, ylim=c(0, 1), freq=FALSE)
lines(density(d$O.ring), col='blue', lwd=2)
```


# Code 1C: Univariate analysis on Temp

```{r}
# kernel density of launch temperature
plot(density(d$Temp), main='Kernel Density of Launch Temperature', xlab='Temperature (F)')
rug(d$Temp)
# boxplot of launch temperature
boxplot(d$Temp, main='Boxplot of Launch Temperature', xlab='Temperature (F)')
```

# Code 2A: plot figure 1, joint temperature versus number of O-rings failure incidents

##  Using plot

```{r}
# average temperature for flights with at least one O-rings failure
avg.temp.failure <- d %>% filter(., O.ring > 0) %>% .$Temp %>% mean(.)
# average temperature for flights with no O-rings failure
avg.temp.success <- d %>% filter(., O.ring == 0) %>% .$Temp %>% mean(.)
# average temperature for all flights
avg.temp <- mean(d$Temp)
# plot joint temperature vs number of O-rings failure incidents
plot(main='Joint Temperature vs Number of O-rings Failure Incidents',
     x=d$Temp, xlab='Temperature (F)', xlim=c(xlim.min,xlim.max),
     y=d$O.ring, ylab='Number of O-rings Failure', yaxt='n', 
     panel.first=grid(col="gray", lty="dotted"), pch=20)
abline(h=0, v=c(avg.temp.failure, avg.temp.success, avg.temp, temp.launch), 
       col=c('black', 'red', 'green', 'gray', 'blue'))
axis(side=2, at=0:2)
```

## Using ggplot (TODO: add abline at 0, format y-axis to show whole numbers, add avg temps)

```{r}
ggplot(d, aes(x=Temp, y=O.ring)) + 
  geom_point() + 
  labs(title = 'Joint Temperature vs Number of O-Rings Failure Incidents', 
       x='Temperature (F)', y='Number of O-rings Failure') 
# + theme(plot.title = element_text(lineheight=1, face="bold", color = "darkred"))
```

# Code 3: (TODO)

```{r}
unique(d$Pressure)
```


```{r}
d %>%
  ggplot(aes(x = factor(O.ring), fill = factor(O.ring))) +
  geom_bar() +
  facet_wrap(~Pressure) + 
  labs(title = 'Count of O-Rings Failure at Different Pressure Levels', y='Count', x='Number of O-rings failure') + 
  theme(plot.title = element_text(lineheight=1, face="bold", color = "lightseagreen")) 
```

```{r}
#ggplot(d, aes(x=Pressure, y=O.ring)) + geom_point() + 
#  labs(title = 'Joint Pressure Versus Number of O-Ring Failures', y='Number of O-ring failures', #x='Pressure (psi') + 
#  theme(plot.title = element_text(lineheight=1, face="bold", color = "lightseagreen")) 
```

# Code 4A: linear regression model 

```{r}
lm1 <- lm(O.ring ~ Temp, data=d)
summary(lm1)
```


The ordinary least squares, simple linear regression model, prediction equation is given by 

predicted damage = 3.69841 - 0.04754 * Temp 


```{r}
predict(lm1, data.frame(Temp = 31), type='response')
```

The predicted damage can be interpreted as the predicted probability of damage for the given temperature.
Plugging various temperatures into this equation we get somewhat reasonable predicted probabilities.
For a temperature of 77, the predicted probability of damage is 0.025. For a temperature of 65,
the predicted probability is 0.474. Finally, for a temperature of 51, the predicted probability is 0.998.
As temperature decreases the predicted probability of damage increases. Figure 3.5 shows the fit of the
simple linear regression model.

Although this appears to be a reasonable prediction equation, it can lead to nonsensical predictions.
For example, if the temperature is 85, the predicted probability of damage is -0.274. This is nonsense
since probabilities can never be less than zero. Similarly, any temperature 50 or below will give a
predicted probability of more than one. Again, this is nonsense. The bounded nature of probabilities is
not incorporated into the simple linear regression model.

Another clue to the inadequacy of the simple linear model for the Challenger data is shown in the
pattern of the residuals. Rather than exhibit a random scatter of points, the plot of residuals versus
temperature (see Figure 3.6) shows two slanted lines.


```{r}

lm1.residuals <- residuals(lm1)
plot(d$Temp, lm1.residuals, xlab="temperature", ylab="Residuals:SLR model", main="Plot of residuals (SLR model)")
```

When ordinary least squares is used to fit a simple linear regression model, the estimates of the
intercept and slope will be unbiased. That is, on average the intercept and slope estimates give the
correct values. Since we are violating the equal variance assumption (recall the variance of a binary
response will change with changes in the mean), the standard errors for the estimates of intercept and
slope will be larger than they should be. Additionally, the unusual pattern in the residuals indicates
that the simple linear model does not provide a good fit to the data.
When variances are unequal, weighted least squares (an extension of ordinary least squares) can be
used. In essence, observations with large variances are given lower weights and thus have less influence
on the fit of the simple linear model. Ordinarily the weights are inversely proportional to the variances.
That is



# Code 5A: binary logistic regression model 

A better approach is to fit the linear regression model to a function of π that is
bounded by 0 and 1 and with which the binomial random variable at least theoretically
has a linear relationship. Two such functions are the logit , defined as 

log(pi/(1-pi))


# Code 5B: hypothesis of binary logistic regression model 

# Code 5C: fitting the model

```{r}
n <- sum(d$Number)
logm1 <- glm(O.ring/Number ~ Temp + Pressure, weights=Number, family=binomial, data=d)
summary(logm1)
deviance(logm1) + 2*3*(n/(n - 3 - 1))
```

# Code 6A: binomial model 

We apply the logistic regression model to model the probability of the O-rings failure at various temperatures.

$$\begin{aligned} 
P(Y_i = 1 | x_i) = \frac{e^{\beta_0 + \beta_1 x_i}}{1 + e^{\beta_0 + \beta_1 x_i}} = \pi_i 
\end{aligned}$$

$i$: observation for a single launch
$Y_i = 1$: failure launch, at least one O-ring failure
$Y_i = 0$: successful launch, no O-ring failure
$x_i$: outside temperature in degrees Fahrenheit

# Code 6B: hypothesis of binomial model 

$H_0: \beta_1 = 0$

$H_a: \beta_1 \ne 0$

Null Hypothesis: outside temperature has no effect in O-rings failure

Alternative Hypothesis: outside temperature has an effect in O-rings failure

# Code 6C: fitting the model

```{r}
bm0 <- glm(formula = O.ring > 0 ~ 1, family=binomial,  data=d)
bm1 <- glm(formula = O.ring > 0 ~ Temp, family=binomial, data=d)
bm2 <- glm(formula = O.ring > 0 ~ Pressure, family=binomial, data=d)
bm3 <- glm(formula = O.ring > 0 ~ Temp + Pressure, family=binomial, data=d)
stargazer(bm0, bm1, bm2, bm3, header=FALSE, type='text')
bm1.b0 <- bm1$coefficients[1]
bm1.b1 <- bm1$coefficients[2]
```

$$\begin{aligned}
P(Y_i = 1 | x_i) & = \frac{e^{\beta_0 + \beta_1 x_i}}{1 + e^{\beta_0 + \beta_1 x_i}} \\
                 & \approx \frac{e^{15.0429 - 0.232 x_i}}{1 + e^{15.0429 - 0.232 x_i}} 
                 = \hat{\pi_i}
\end{aligned}$$

# Code 6D: Likelihood Ratio Test (LRT) to determine if Temp was a statistically significant addition to model

```{r}
# likelihood ratio test
# method #1
pchisq(-2 * (logLik(bm0) - logLik(bm1)), df=1, lower.tail=FALSE)
# method #2 (recommended)
anova(bm0, bm1, bm2, test='Chi')
```

# Code 6E: expected value and standard error

$$\begin{aligned}
E(Y_i) = 1 \cdot \pi_i + 0 \cdot (1 - \pi_i) = \pi_i
\end{aligned}$$


$$\begin{aligned}
\sigma^2_{y_i} & = E(Y_i - E(Y_i))^2 \\
               & = E(Y_i)(1 - E(Y_i)) \\
               & = \pi_i (1 - \pi_i)
\end{aligned}$$


# Code 6F: concluding remakrs about binomial model 

## p-value

With p-value = 0.0320, there is enough evidence to reject the null hypothesis and conclude that outside temperature has an effect in O-ring failure.

## Interpret the main result of your final model in terms of both odds and probability of failure.

### 1. odds

$$\begin{aligned}
odds(x_i) & = \frac{\pi_i}{1 - \pi_i} = \frac{\frac{e^{\beta_0 + \beta_1 x_i}}{1 + e^{\beta_0 + \beta_1 x_i}}}{1 - \frac{e^{\beta_0 + \beta_1 x_i}}{1 + e^{\beta_0 + \beta_1 x_i}}} \\ 
          & = e^{\beta_0 + \beta_1 x_i}
\end{aligned}$$


```{r}
# for every 1 F increase, odds of the o-rings failing decreases by a factor of 
exp(bm1.b1)
# for every 1 F decrease, odds of the o-ring failing increases by a factor of 
exp(-bm1.b1)
```

For every 1 degree Fahrenheit increase in temperature, the odds of the O-rings failing for a given launch decreases by a factor of 0.79. 

For every 1 degree Fahrenheit decrease in temperature, the odds of the O-rings failing for a given launch increases by a factor of 1.26. 

```{r}
# for every 1 F increase, odds of the o-rings failing decreases by ... % 
100 * (exp(bm1.b1) - 1)
# for every 1 F decrease, odds of the o-rings failing increases by ... % 
100 * (exp(-bm1.b1) - 1)
```

For every 1 degree Fahrenheit increase in temperature, the odds of the O-rings failing for a given launch decreases by 20%. 

For every 1 degree Fahrenheit decrease in temperature, the odds of the O-rings failing for a given launch increases by 26%. 

### 2. odds ratio

$\text{odds ratio} = \frac{odds(x_i + 1)}{odds(x_i)} = \frac{e^{\beta_0 + \beta_1(x_i + 1)}}{e^{\beta_0 + \beta_1(x_i)}} = e^{\beta_1}$

$\hat{\text{odds ratio}} = e^{\hat{\beta_1}} = e^{-0.2321627} = 0.7928171$

```{r}
# estimated odds ratio
exp(bm1.b1)
```

95% CI for estimated odds ratio: $(e^{\hat{\beta_1} - 1.96 \cdot SE(\hat{\beta_1})}}, e^{\hat{\beta_1} + 1.96 \cdot SE(\hat{\beta_1})}})$

```{r}
# 95% ci for odds ratio
get.glm.se <- function(model) sqrt(diag(vcov(model)))
c(exp(bm1.b1 - 1.96 * get.glm.se(bm1)['Temp']), exp(bm1.b1 + 1.96 * get.glm.se(bm1)['Temp']))
```

The odds of failure decreases as temperature increases.

### 3. probability of failure

Since the Challenger shuttle was launched at 31 F, the proability of O-rings failing based on our logistic regression model is approximately $\hat{\pi_i} \approx 0.99.$

```{r}
predict(bm1, data.frame(Temp=temp.launch), type='response')
```

If they have waited until 53 F as suggested, the probability of O-rings failing based on our logistic regression model is approximately 0.94. 

```{r}
predict(bm1, data.frame(Temp=53), type='response')
```

This would have decreased the odds of a failure by a factor of 0.006. 

```{r}
# If launched at 53 F instead of 31, it would have decreased odds of failure by a factor of
exp(bm1.b1  * (53 - temp.launch))
```

The probability of all five O-rings failing is 99.8%, which is highly likely.

```{r}
predict(bm1, data.frame(Temp=temp.launch), type='response')^5
```

# Code 2B: plotting figure 4, joint temperature versus number of O-rings failure incidents with linear, binary and binomial model

## Using plot

```{r}
plot(main='Joint Temperature vs Number of O-rings Failure Incidents',
     x=d$Temp, xlab='Temperature (F)', xlim=c(xlim.min,xlim.max),
     y=d$O.ring, ylab='Number of O-rings Failure', yaxt='n', 
     panel.first=grid(col="gray", lty="dotted"), pch=20)
abline(h=0)
axis(side=2, at=0:2)

# linear model
abline(lm1, col='red')
# add confidence interval
get.predicted.linear.ci <- function(model, newdata, alpha=0.05) {
  y.hat <- predict(model, newdata, interval='predict')
  list(lower=y.hat[,'lwr'], upper=y.hat[,'upr'])
}
input <- seq(xlim.min, xlim.max)
ci <- get.predicted.linear.ci(lm1, data.frame(Temp=input))
lines(input, ci$lower, col='red', lty='dotdash')
lines(input, ci$upper, col='red', lty='dotdash')

# binary logistic regression model: TODO

# binomial model: method 1
curve(exp(bm1.b0 + bm1.b1*x)/(1 + exp(bm1.b0 + bm1.b1*x)), 
      from=xlim.min, to=xlim.max, add=TRUE, col='blue')
# binomial model: method 2
curve(predict.glm(bm1, data.frame(Temp=x), type='response'), 
      from=xlim.min, to=xlim.max, add=TRUE, col='blue')
# add confidence interval 
get.predicted.binomial.ci <- function(model, newdata, alpha=0.05) {
  y.hat <- predict.glm(model, newdata, type='link', se.fit=TRUE)
  y.hat.lower <- y.hat$fit - qnorm(1 - alpha/2) * y.hat$se.fit
  y.hat.upper <- y.hat$fit + qnorm(1 - alpha/2) * y.hat$se.fit
  pi.hat.lower <- exp(y.hat.lower) / (1 + exp(y.hat.lower))
  pi.hat.upper <- exp(y.hat.upper) / (1 + exp(y.hat.upper))
  list(lower=pi.hat.lower, upper=pi.hat.upper)
}
curve(get.predicted.binomial.ci(bm1, data.frame(Temp=x))$lower, 
      add=TRUE, col='blue', lty='dotdash')
curve(get.predicted.binomial.ci(bm1, data.frame(Temp=x))$upper, 
      add=TRUE, col='blue', lty='dotdash')
```


# Code 8: predicted confidence interval with temperature at launch

## with binomial model 

```{r}
get.predicted.binomial.ci(bm1, data.frame(Temp=temp.launch))
```

# Code 9: bootstrap for 5e

parametric bootstrap to compute intervals:

1. simulate large number of data sets (n = 23 for each) from the estimated model of $logit(\pi) = \beta_0 + \beta_1 cdot Temp$

2. estimate new models for each data set, say $logit(\hat{\pi}^{*}) = \hat{\beta_0^{*} + \hat{\beta_1^{*} \cdot Temp$

3. compute $\hat{\pi}^{*}$ at a specific temperature of interest.

The author used the 0.05 and 0.95 observed quantiles from the $\hat{\pi}^{*}$ simulated distribution as their 90% confidence interval limits. 

Using the parameteric bootstrap, compute 90% confidence intervals separately at temperature 31 and 72.


My understanding is there should be two stages of randomization.  1) randomize the `x` variable (=temp); then 2) randomize the binomial output of all of the n (=23) trials in each sample (=fraction failed).  My assumption was we should put both (1) and (2) inside the for 1=1:10000 loops.  

But in Bilder's example (http://www.chrisbilder.com/categorical/Chapter2/SamplingDist.R, under the comment "Simulate more than one data set and calculate true confidence level ") it appears he does (1) outside of the loop, and (2) inside.  He also doesn't claim it's a "bootstrap"; he just calls it a Monte Carlo simulation.  

First, this happens before `sim.all` is called:
```x<-runif(n = sample.size, min = 0, max = 1)
  pi<-exp(beta0 + beta1*x) / (1 + exp(beta0 + beta1*x))```
Then he calls sim.all (he has 3 versions in the code, 2 are commented out by default):
   ```save.it<-sim.all(sample.size = sample.size, number.data.sets = 10000, x = x, seed.numb = 1829)```
 Inside sim.all, he calculates the response variable for all the simulations at once:
```y<-rbinom(n = length(x)*number.data.sets, size = 1, prob = pi)```
and rearranges these values into a matrix.  Then he calls his `calc2` function to build a new glm model for each simulation (=column of the matrix) via `apply`.  `apply` iterates through the columns of the matrix to get y values, but pairs each with the fixed values of the same x vector:
```save.results<-apply(X = y.mat, MARGIN = 2, FUN = calc2, x = x)```
(edited)

elegant, but once he's picked his `x` values (randomly) once, they are used in all 10,000 runs

```{r}
library(dplyr)
sample_n(d, sample.size, replace=TRUE)
```


```{r}
library(dplyr)

d <- read.csv("./dataset/challenger.csv", header=TRUE, sep=",")

# set seed number to reproduce results
set.seed(101)

bootstrap <- function(sample.size, total.O.rings, pred.temp, original.data){
  # simulate temperature at launch
  data <- sample_n(original.data, sample.size, replace=TRUE)
  newdata <- data.frame(Temp=data$Temp)
  
  # find pi.hat 
  m <- glm(formula=O.ring/Number ~ Temp, weights=Number, family=binomial, data=data)
  pi.hat <- predict(object=m, newdata=newdata, type='response')

  # simulate number of O-rings failure
  newdata$O.ring <- rbinom(n=sample.size, size=total.O.rings, pi.hat)
  newdata$Number <- replicate(sample.size, total.O.rings)
  
  # estimated model 
  estimated.m <- glm(formula=O.ring/Number ~ Temp, weights=Number, family=binomial(link=logit), data=newdata)
  
  # make predictions
  estimated.pi.hat <- predict(object=estimated.m, newdata=data.frame(Temp=pred.temp), type='response')

  return(estimated.pi.hat)
}

# provide the variables to pass in to bootstrap
# 1. sample size
sample.size <- 23 
#sample.size <- nrow(d)

# 2. total number of O-rings
total.O.rings <- 6
#total.O.rings <- unique(d$Number)

# 3. provide predicted temperature
pred.temp <- 31

# 5. run 10000 times
estimated.pi.hats <- replicate(10000, bootstrap(sample.size, total.O.rings, pred.temp, d))
quantile(estimated.pi.hats, c(0.05, 0.95))
```



```{r}
library(dplyr)

# remove all objects from current workspace
rm(list = ls())

# load dataset
d <- read.csv("./dataset/challenger.csv", header=TRUE, sep=",")

# set seed number to reproduce results
set.seed(101)

bootstrap <- function(sample.size, total.O.rings, pred.temp, original.data){
  
  Temp <- sample(d$Temp, sample.size)
  pi.hat <- predict(object=m, newdata=data.frame(Temp=Temp), type='response')
  O.ring <- rbinom(n=sample.size, size=total.O.rings, pi.hat)
  Number <- replicate(sample.size, total.O.rings)
  
  newdata <- data.frame(Temp = Temp,
                        O.ring = O.ring,
                        Number = Number)
  
  # estimated model 
  estimated.m <- glm(formula=O.ring/Number ~ Temp, weights=Number, family=binomial(link=logit), data=newdata)
  
  # make predictions
  estimated.pi.hat <- predict(object=estimated.m, newdata=data.frame(Temp=pred.temp), type='response')
  
  return(estimated.pi.hat)
}

# provide the variables to pass in to bootstrap
# 1. sample size
sample.size <- 23 
#sample.size <- nrow(d)

# 2. total number of O-rings
total.O.rings <- 6
#total.O.rings <- unique(d$Number)

# 3. provide predicted temperature
pred.temp <- 31

# 4. provide the model
m <- glm(formula=O.ring/Number ~ Temp, weights=Number, family=binomial, data=d)

# 5. run 10000 times
estimated.pi.hats <- replicate(10000, bootstrap(sample.size, total.O.rings, pred.temp, d))
quantile(estimated.pi.hats, c(0.05, 0.95))
```

```{r}
bootstrap_coef <- function(temp_data, model, pred_temp) {
  # 1. simulate large number of datasets from estimated model
  sample_temps <- sample(x = temp_data, size = 23, replace = TRUE)
  sample_preds <- predict(object = model, newdata = data.frame(Temp = sample_temps), type = "response")

  #build a dataframe
  df <- data.frame(temp = sample_temp, 
             prob_pred = sample_preds
             )
  
  #map function to find predicted class (fail or not fail)
  df$fail_pred <- unlist(lapply(X = df$prob_pred, FUN = function(x) rbinom(1, 1, x)))
  
  #build glm model
  bootstrap_glm <- glm(formula = fail_pred ~ temp, family = binomial(link=logit), data = df)
  
  #predict probability at pred_temp
  predict(object = bootstrap_glm, newdata = data.frame(temp = pred_temp), type = "response")
}

est_prob_31 <- replicate(10000, bootstrap_coef(d$Temp, bm1, 31), simplify = FALSE)

est_prob_72 <- replicate(10000, bootstrap_coef(d$Temp, bm1, 72), simplify = FALSE)

quantile(unlist(est_prob_31), c(.05, .95))
quantile(unlist(est_prob_72), c(.05, .95))
```

```{r}
library(boot)
R <- 10000
# obtain r-square
rsq <- function(formula, data, indices) {
  sample <- data[indices,]
  fit <- lm(formula, data=sample)
  return(summary(fit)$r.square)
}
# bootstrapping with 10000 replications
results <- boot(data=d, statistic=rsq, R=R, formula=O.ring > 0 ~ Temp)
results
plot(results)
boot.ci(results, type='bca')
```

# Code 10: quadratic term to model temperature

```{r}
qm <- glm(O.ring > 0 ~ Temp + I(Temp^2), family=binomial(link=logit), data=d) 
summary(qm1)
anova(bm1, qm, test='Chisq')
```

This quadratic model is not necessary. 

